Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models

### 2020-09-15 analysis/paper/paper.Rmd
made some formatting changes and minor edits for submission to Ecosphere.
made some revisions based on reviewer #2's comments

A number of studies have used other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O. clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are usually unavailable [@Cramer2009].

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook salmon parr, there is clearly movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population levels. This is an efficient technique to leverage existing relationships for meaningful management decisions.

@@

One of the potential downsides to the extrapolation approach used here is that the GAAs generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonina2019].

@@

```{r gaa-covars, eval = T}
tibble(Covariate = gaa_covars) %>%
  left_join(gaa_meta %>%
              select(Covariate = FieldName,
                     Name = DisplayName,
                     Units,
                     SpatialScale)) %>%
  mutate(Units = if_else(Units == "na",
                         as.character(NA),
                         Units),
         Units = if_else(grepl('\\^2', Units),
                         str_replace(Units, '\\^2', '$^2$'),
                         Units)) %>%
  mutate(Name = if_else(Covariate == 'S2_02_11' & is.na(Name),
                        'NorWeST Aug. Temperature',
                        Name),
         Units = if_else(Covariate == 'S2_02_11',
                                'C',
                                Units),
         SpatialScale = if_else(Covariate == 'S2_02_11',
                                'Reach-2km',
                                SpatialScale)) %>%
  left_join(model_svy_df$mod_full[[1]] %>%
              summary() %>%
              coefficients() %>%
              as_tibble(rownames = "Covariate") %>%
              select(1:3)) %>%
  select(-Covariate) %>%
  rename(`Spatial Scale` = SpatialScale,
         Covariate = Name) %>%
  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
        booktabs = T, 
        escape = F,
        digits = 3) %>%
  kable_styling()

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T) %>%
  kable_styling() %>%
  # kable_styling(latex_options = c('scale_down')) %>%
  column_spec(1, width = "1in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile regression forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  mutate(imp = (imp / max(imp)) * 100) %>%
  arrange(desc(imp)) %>%
  left_join(hab_dict %>%
              filter(is.na(MetricGroupName) | MetricGroupName != 'Tier 1 Summary')) %>%
  select(-Name) %>%
  left_join(sel_hab_mets %>%
              select(ShortName = Metric,
                     Name)) %>%
  mutate_at(vars(ShortName, Name, MetricCategory),
            list(~ fct_reorder(., .x = imp))) %>%
  ggplot(aes(x = Name,
             y = imp)) +
  # geom_col(aes(fill = MetricCategory)) +
  # scale_fill_brewer(palette = 'Set2') +
  geom_col(fill = 'gray40') +
  coord_flip() +
  labs(x = 'Metric',
       y = 'Relative Importance')

@@

```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values. Tick marks along the X-axis depict observed values."}
plot_partial_dependence(rf_mod = qrf_mod,
                        data = qrf_data,
                        data_dict = hab_dict %>%
                          select(-Name) %>%
                          inner_join(sel_hab_mets %>%
                                       select(ShortName = Metric,
                                              Name)),
                        type = 'quantile',
                        pred_quantile = 0.9,
                        log_transform = T,
                        log_offset = dens_offset,
                        scales = 'free') +
  labs(y = 'Capacity Prediction (per m)') +
  scale_color_grey(start = 0.5, end = 0.5) +
  theme(legend.position = 'none',
        strip.text = element_text(size = 6))
```

>>>>

A number of studies have used other modeling approaches to elicit fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O. clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are unavailable in most watersheds [@Cramer2009].

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor proxy of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile random forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. In addition, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watersheds within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].  We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the upstream limit of master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists, and the downstream limit by when streams were no longer wadeable (often by some combination of estimated bankfull width and cumulative drainage area). The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook salmon parr, our fish data clearly shows movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population levels. This is an efficient technique to leverage existing relationships for meaningful management decisions.

@@

One of the potential downsides to the extrapolation approach used here is that the GAAs generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) may be on the horizon [@Tonina2019].

@@

```{r gaa-covars, eval = T}
tibble(Covariate = gaa_covars) %>%
  left_join(gaa_meta %>%
              select(Covariate = FieldName,
                     Name = DisplayName,
                     Units,
                     SpatialScale)) %>%
  mutate(Units = if_else(Units == "na",
                         as.character(NA),
                         Units),
         Units = if_else(grepl('\\^2', Units),
                         str_replace(Units, '\\^2', '$^2$'),
                         Units)) %>%
  mutate(Name = if_else(Covariate == 'S2_02_11' & is.na(Name),
                        'NorWeST Aug. Temperature',
                        Name),
         Units = if_else(Covariate == 'S2_02_11',
                                'C',
                                Units),
         SpatialScale = if_else(Covariate == 'S2_02_11',
                                'Reach-2km',
                                SpatialScale)) %>%
  left_join(model_svy_df$mod_full[[1]] %>%
              summary() %>%
              coefficients() %>%
              as_tibble(rownames = "Covariate") %>%
              select(1:3)) %>%
  select(-Covariate) %>%
  rename(`Spatial Scale` = SpatialScale,
         Covariate = Name) %>%
  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile random forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
        booktabs = T, 
        escape = F,
        digits = 3) %>%
  kable_styling()

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T) %>%
  kable_styling() %>%
  # kable_styling(latex_options = c('scale_down')) %>%
  column_spec(1, width = "1in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  mutate(imp = (imp / max(imp)) * 100) %>%
  arrange(desc(imp)) %>%
  left_join(hab_dict %>%
              filter(is.na(MetricGroupName) | MetricGroupName != 'Tier 1 Summary')) %>%
  select(-Name) %>%
  left_join(sel_hab_mets %>%
              select(ShortName = Metric,
                     Name)) %>%
  mutate_at(vars(ShortName, Name, MetricCategory),
            list(~ fct_reorder(., .x = imp))) %>%
  ggplot(aes(x = Name,
             y = imp)) +
  # geom_col(aes(fill = MetricCategory)) +
  # scale_fill_brewer(palette = 'Set2') +
  geom_col(fill = 'gray40') +
  coord_flip() +
  labs(x = 'Metric',
       y = 'Relative Importance')

@@

```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values. Tick marks along the X-axis depict observed values."}
plot_partial_dependence(rf_mod = qrf_mod,
                        data = qrf_data,
                        data_dict = hab_dict %>%
                          select(-Name) %>%
                          inner_join(sel_hab_mets %>%
                                       select(ShortName = Metric,
                                              Name)),
                        type = 'quantile',
                        pred_quantile = 0.9,
                        log_transform = T,
                        log_offset = dens_offset,
                        scales = 'free') +
  labs(y = 'Capacity Prediction (per m)') +
  scale_color_grey(start = 0.5, end = 0.5) +
  theme(legend.position = 'none',
        strip.text = element_text(size = 6))
```
